Marker_identification_Benchmark

### 2023-11-03 Marker_identfication_benchmark.Rmd
path adjustment

---
title: "Marker_identification_Benchmark"
output: pdf_document
date: '2023-10-16'
---
# Load Packages
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("BiocManager", quietly = TRUE)) BiocManager::install("Seurat")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

@@

# Load Data & Process
## Load Data
```{r}
path <- "/data/ruiqi/local_marker/dataset/tabular_muris/"
tissue_name = "marrow_facs" #lung_facs, pancreas_facs
tiss <- readRDS(paste0(path,tissue_name,".rds"))
```
## Process
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class

@@

# Define Ground truth Marker
## Criterion1: Fold-change
For each gene, sorting cell types by mean expression for each gene and computing fold changes between consecutive types, take the maximum value among the N-1 fold change values, given N cell types.
```{r}
path = "./benchmark_result/"
tissue_name = "marrow_facs"
```

@@

max_logfc = sort(max_logfc,decreasing = TRUE)
write.table(max_logfc,file = paste0(path,tissue_name,"_ground_truth_c1.txt"))
```
## Criterion2: CellMarkerDB
Downloaded marker from 
http://117.50.127.228/CellMarker/CellMarker_download_files/file/Cell_marker_Seq.xlsx
<!-- PanglaoDB: https://panglaodb.se/markers/PanglaoDB_markers_27_Mar_2020.tsv.gz and exclude genes that reported only once -->
```{r}
# Load CellMarkerDB
cell_marker_db = readxl::read_xlsx("/data/ruiqi/local_marker/dataset/Cell_Marker/Cell_marker_All.xlsx") %>% filter(species == "Mouse") %>% filter(tissue_class == "Bone marrow") %>% filter(!is.na(cellontology_id)) %>% filter(!is.na(PMID)) %>% mutate(gene_name1 = marker) %>% mutate(gene_name2 = Symbol) %>% select(c("gene_name1","gene_name2")) 
# %>% group_by(dplyr::across(dplyr::everything())) %>% dplyr::summarise(count = n(), .groups = "drop") %>% filter(count >= 1) # Lung, Pancreas

# Match gene name
celldb_marker = union(cell_marker_db$gene_name1,cell_marker_db$gene_name2)
celldb_marker = unlist( lapply(celldb_marker,function(i){
  id = grep(paste0("^", i, "$"), selected_genes, ignore.case = TRUE)
  if(length(id)==0){return(NA)}
  else{return(selected_genes[id])}
}) )
celldb_marker = celldb_marker[!is.na(celldb_marker)] %>% unique()
write.table(celldb_marker,file = paste0(path,tissue_name,"_ground_truth_c2.txt"))
```

# Run Each Method and save results
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```

## LocalMarkerDetector
Genes were ranked by the increasing order of Cumulative Diffusion-KL score
```{r}
method = "lmd"
source("./LMD_function.R", echo = F)
res = LMD(dat,feature_space,max_time = 2^10,knn = 7)
marker = show_result_lmd(res, n = length(selected_genes))
marker[order(marker$score),'rank'] = 1:nrow(marker)
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```

## Seurat v3 Wilcoxon unfiltered
Genes were ranked by their lowest adjusted P-value in any cluster (increasing order), if two genes had the same, prioritize genes with larger FC.
```{r}
tiss <- FindNeighbors(tiss, dims = 1:n_dim)
tiss <- FindClusters(tiss, resolution = 1.5, algorithm = 3)
method = "wilcox_no_filter"
marker <- FindAllMarkers(tiss, features = selected_genes, only.pos = FALSE, test.use = "wilcox", logfc.threshold = 0, min.pct = 0)
gene_rank = marker %>% arrange(.,p_val_adj,desc(avg_log2FC)) %>% distinct(.,gene, .keep_all = TRUE) %>% select(gene)
gene_rank$'rank' = 1:nrow(gene_rank)
marker <- merge(marker,gene_rank,by = "gene")
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```
## Seurat v3 default
Genes were ranked by their lowest adjusted P-value in any cluster (increasing order), if two genes had the same, prioritize genes with larger FC.
```{r}
tiss <- FindNeighbors(tiss, dims = 1:n_dim)
tiss <- FindClusters(tiss, resolution = 1.5, algorithm = 3)

@@

gene_rank = marker %>% arrange(.,p_val_adj,desc(avg_log2FC)) %>% distinct(.,gene, .keep_all = TRUE) %>% select(gene)
gene_rank$'rank' = 1:nrow(gene_rank)
marker <- merge(marker,gene_rank,by = "gene")
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```
## singleCellHaystack
https://alexisvdb.github.io/singleCellHaystack/articles/examples/a02_example_scRNAseq.html
Genes were ranked based on the increasing order of log(adjusted P-value)
```{r}
method = "haystack"
library(singleCellHaystack)
set.seed(123)
res <- haystack(x = feature_space, expression = dat)
marker <- show_result_haystack(res.haystack = res)
marker[order(marker$log.p.adj),"rank"] = 1:nrow(marker)

write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```
## SEMITONES
https://github.com/ohlerlab/SEMITONES/tree/master
https://github.com/ohlerlab/SEMITONES_paper/blob/main/notebooks/10_SEMITONES_benchmark_corrected.ipynb

@@

write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```

@@

# Determining informative genes
hs_results = hs.compute_autocorrelations()
```
```{r}
method = "hotspot"
marker = reticulate::py$hs_results
marker[order(marker$FDR),"rank"] = 1:nrow(marker)
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```

@@

result = MarcoPolo.find_markers(adata=adata, regression_result=regression_result)
```
```{r}
method = "marcopolo"
marker = reticulate::py$result
marker$'rank' = marker$MarcoPolo_rank + 1
write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = TRUE)
```

@@

write.table(marker,file = sprintf("%s%s_%s.csv",path,method,tissue_name),row.names = FALSE)
```

# Integrate rank from all methods into one dataframe
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```
```{r}
method_ls = c("lmd","haystack","hotspot","marcopolo","semi","wilcox_default","wilcox_no_filter","hvg")
method_fullname = setNames(c("LMD","singleCellHaystack","Hotspot","Marcopolo","SEMITONES","Seurat v3 default", "Seurat v3 Wilcoxon unfiltered","HighlyVariableGenes"),method_ls)

df_benchmark = do.call(cbind,lapply(method_ls, function(method){
  marker = read.table(sprintf("%s%s_%s.csv",path,method,tissue_name),header = TRUE)
  if("gene" %in% colnames(marker)){
    return(marker$rank[match(selected_genes,marker$gene)])
  }else{
    return(marker$rank[match(selected_genes,rownames(marker))])
  }
}))
df_benchmark[is.na(df_benchmark)] = nrow(df_benchmark)
df_benchmark = data.frame(df_benchmark,row.names = selected_genes)
colnames(df_benchmark) = method_ls

write.table(df_benchmark,file = paste0(path,tissue_name,"_benchmark_rank_table.csv"),row.names = TRUE)
```

# AUROC
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
df_benchmark = read.table(paste0(path,tissue_name,"_benchmark_rank_table.csv"))

method_ls = c("lmd","haystack","hotspot","marcopolo","semi","wilcox_default","wilcox_no_filter","hvg")
method_fullname = setNames(c("LMD","singleCellHaystack","Hotspot","Marcopolo","SEMITONES","Seurat v3 default", "Seurat v3 Wilcoxon unfiltered","HighlyVariableGenes"),method_ls)
```
```{r}
max_logfc = read.table(paste0(path,tissue_name,"_ground_truth_c1.txt")) %>% rownames()
celldb_marker = read.table(paste0(path,tissue_name,"_ground_truth_c2.txt"))[,1]
gt_list = c(lapply(seq(50,1000,50),function(x) max_logfc[1:x]),list(celldb_marker))
names(gt_list) = c(paste0("Top",seq(50,1000,50)),"CellMarkerDB")

@@

  df = roc_df %>% select(c("Method","AUROC")) %>% distinct()
  df$'gt_set' = names(gt_list)[i]
  df
}))
df$Method = as.factor(df$Method)
levels(df$Method) = method_fullname[levels(df$Method)]
df$gt_set = factor(df$gt_set,levels = names(gt_list))
write.table(df,file = paste0(path,tissue_name,"_auroc.csv"),row.names = FALSE)
```

@@

# Density Index
```{r}
path = "./benchmark_result/"
# tissue_name = "marrow_facs"
```
```{r}
df_benchmark = read.table(paste0(path,tissue_name,"_benchmark_rank_table.csv"))

@@

write.table(score_measure,file = paste0(path,tissue_name,"_density_index.csv"),row.names = FALSE)
```

# Sensitivity Test (Try different K)
```{r}
path = "./marrow_result/"
# tissue_name = "marrow_facs"
```
```{r}
# RunLMD under differen K
source("./LMD_function.R", echo = F)
score_df = do.call(cbind,lapply(c(3:10,20,30,40,50),function(kNN){
  try({
    res = LMD(dat,feature_space,knn = kNN)
    return(res$cumulative_score)
  },silent = TRUE)
}))
colnames(score_df) = paste0("kNN",c(3:10,20,30,40,50))
write.table(score_df,file = paste0(path,tissue_name,"_different_knn.csv"),row.names = TRUE)
```
```{r}
# Obtain AUROC
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))
path = "./benchmark_result/"
max_logfc = read.table(paste0(path,tissue_name,"_ground_truth_c1.txt")) %>% rownames()
celldb_marker = read.table(paste0(path,tissue_name,"_ground_truth_c2.txt"))[,1]

@@

auc_df = apply(score_ranked,2,function(x){
  unlist(lapply(gt_list,function(genes){
    roc = pROC::roc(as.integer(rownames(score_ranked) %in% genes), x, direction = ">")
    pROC::auc(roc)
  }))
})
path = "./marrow_result/"
auc_df = data.frame(auc_df,gt = names(gt_list))
write.table(auc_df,file = paste0(path,tissue_name,"_sensitivity_auroc.csv"),row.names = FALSE)
```

>>>>

---
title: "Marker_identification_Benchmark"
author: "Ruiqi Li"
output: pdf_document
date: '2023-10-16'
---
# Load Packages
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("BiocManager", quietly = TRUE)) BiocManager::install("Seurat")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

@@

# Load Data & Process
## Load Data
```{r}
tissue_fullname = "Marrow"

if(tissue_fullname == "Bone marrow"){
  tissue_name = "marrow_facs"
  if (!file.exists('facs_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000)
  download.file("https://figshare.com/ndownloader/files/13092380",
                destfile = 'facs_Marrow_seurat_tiss.Robj', method = 'libcurl')
  }
  load('facs_Marrow_seurat_tiss.Robj')
  tiss <- UpdateSeuratObject(tiss)
}if(tissue_fullname == "Pancreas"){
  tissue_name = "pancreas_facs"
  if (!file.exists('facs_Pancreas_seurat_tiss.Robj')) {
     # If load url doesn't work, please download it and load from your working directory
    options(timeout=6000)
    download.file("https://figshare.com/ndownloader/files/13092386",
                  destfile = 'facs_Pancreas_seurat_tiss.Robj', method = 'libcurl')
  }
  load('facs_Pancreas_seurat_tiss.Robj')
  tiss <- UpdateSeuratObject(tiss)
}if(tissue_fullname == "Lung"){
  tissue_name = "lung_facs"
  if (!file.exists('facs_Lung_seurat_tiss.Robj')) {
     # If load url doesn't work, please download it and load from your working directory
    options(timeout=6000)
    download.file("https://figshare.com/ndownloader/files/13092194",
                  destfile = 'facs_Lung_seurat_tiss.Robj', method = 'libcurl')
   }
  load('facs_Lung_seurat_tiss.Robj')
  tiss <- UpdateSeuratObject(tiss)
}

```
## Process
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class

@@

# Define Ground truth Marker
## Criterion1: Fold-change
For each gene, sorting cell types by mean expression for each gene and computing fold changes between consecutive types, take the maximum value among the N-1 fold change values, given N cell types.
```{r}
folder_path = "benchmark_result"
# Check if the folder exists
if (!dir.exists(folder_path)) {
  # If it doesn't exist, create it
  dir.create(folder_path)
} else {
}
```

@@

max_logfc = sort(max_logfc,decreasing = TRUE)
write.table(max_logfc,file = file.path(folder_path, paste0(tissue_name,"_ground_truth_c1.txt")))
```
## Criterion2: CellMarkerDB
Downloaded marker from 
http://117.50.127.228/CellMarker/CellMarker_download_files/file/Cell_marker_Seq.xlsx
```{r}
# Load CellMarkerDB
cell_marker_db = readxl::read_xlsx("/data/ruiqi/local_marker/dataset/Cell_Marker/Cell_marker_All.xlsx") %>% filter(species == "Mouse") %>% filter(tissue_class == "Bone marrow") %>% filter(!is.na(cellontology_id)) %>% filter(!is.na(PMID)) %>% mutate(gene_name1 = marker) %>% mutate(gene_name2 = Symbol) %>% select(c("gene_name1","gene_name2")) 
# Lung, Pancreas

# Match gene name
celldb_marker = union(cell_marker_db$gene_name1,cell_marker_db$gene_name2)
celldb_marker = unlist( lapply(celldb_marker,function(i){
  id = grep(paste0("^", i, "$"), selected_genes, ignore.case = TRUE)
  if(length(id)==0){return(NA)}
  else{return(selected_genes[id])}
}) )
celldb_marker = celldb_marker[!is.na(celldb_marker)] %>% unique()
write.table(celldb_marker,file = file.path(folder_path, paste0(tissue_name,"_ground_truth_c2.txt")))
```

# Run Each Method and save results
## LocalMarkerDetector
Genes were ranked by the increasing order of Cumulative Diffusion-KL score
```{r}
method = "lmd"
source("LMD_function.R", echo = F)
res = LMD(dat,feature_space,max_time = 2^10,knn = 7)
marker = show_result_lmd(res, n = length(selected_genes))
marker[order(marker$score),'rank'] = 1:nrow(marker)
write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```

## Seurat v4 Wilcoxon unfiltered
Genes were ranked by their lowest adjusted P-value in any cluster (increasing order), if two genes had the same, prioritize genes with larger FC.
```{r}
tiss <- FindNeighbors(tiss, dims = 1:n_dim)
tiss <- FindClusters(tiss, resolution = 1.5, algorithm = 3)
method = "wilcox_no_filter"
marker <- FindAllMarkers(tiss, features = selected_genes, only.pos = FALSE, test.use = "wilcox", logfc.threshold = 0, min.pct = 0)
gene_rank = marker %>% arrange(.,p_val_adj,desc(avg_log2FC)) %>% distinct(.,gene, .keep_all = TRUE) %>% select(gene)
gene_rank$'rank' = 1:nrow(gene_rank)
marker <- merge(marker,gene_rank,by = "gene")
write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```
## Seurat v4 default
Genes were ranked by their lowest adjusted P-value in any cluster (increasing order), if two genes had the same, prioritize genes with larger FC.
```{r}
tiss <- FindNeighbors(tiss, dims = 1:n_dim)
tiss <- FindClusters(tiss, resolution = 1.5, algorithm = 3)

@@

gene_rank = marker %>% arrange(.,p_val_adj,desc(avg_log2FC)) %>% distinct(.,gene, .keep_all = TRUE) %>% select(gene)
gene_rank$'rank' = 1:nrow(gene_rank)
marker <- merge(marker,gene_rank,by = "gene")
write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```
## singleCellHaystack
https://alexisvdb.github.io/singleCellHaystack/articles/examples/a02_example_scRNAseq.html
Genes were ranked based on the increasing order of log(adjusted P-value)
```{r}
method = "haystack"
library(singleCellHaystack)
set.seed(123)
res <- haystack(x = feature_space, expression = dat)
marker <- show_result_haystack(res.haystack = res)
marker[order(marker$log.p.adj),"rank"] = 1:nrow(marker)

write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```
## SEMITONES
https://github.com/ohlerlab/SEMITONES/tree/master
https://github.com/ohlerlab/SEMITONES_paper/blob/main/notebooks/10_SEMITONES_benchmark_corrected.ipynb

@@

write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```

@@

# Determining informative genes
hs_results = hs.compute_autocorrelations()
```
```{r}
method = "hotspot"
marker = reticulate::py$hs_results
marker[order(marker$FDR),"rank"] = 1:nrow(marker)
write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```

@@

result = MarcoPolo.find_markers(adata=adata, regression_result=regression_result)
```
```{r}
method = "marcopolo"
marker = reticulate::py$result
marker$'rank' = marker$MarcoPolo_rank + 1
write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = TRUE)
```

@@

write.table(marker,file = file.path(folder_path, paste0(method,"_",tissue_name,".csv")),row.names = FALSE)
```

# Integrate rank from all methods into one dataframe
```{r}
method_ls = c("lmd","haystack","hotspot","marcopolo","semi","wilcox_default","wilcox_no_filter","hvg")

df_benchmark = do.call(cbind,lapply(method_ls, function(method){
  marker = read.table(file.path(folder_path, paste0(method,"_",tissue_name,".csv")),header = TRUE)
  if("gene" %in% colnames(marker)){
    return(marker$rank[match(selected_genes,marker$gene)])
  }else{
    return(marker$rank[match(selected_genes,rownames(marker))])
  }
}))
df_benchmark[is.na(df_benchmark)] = nrow(df_benchmark)
df_benchmark = data.frame(df_benchmark,row.names = selected_genes)
colnames(df_benchmark) = method_ls

write.table(df_benchmark,file = file.path(folder_path, paste0(tissue_name,"_benchmark_rank_table.csv")),row.names = TRUE)
```

# AUROC
```{r}
df_benchmark = read.table(file.path(folder_path, paste0(tissue_name,"_benchmark_rank_table.csv")))

method_ls = c("lmd","haystack","hotspot","marcopolo","semi","wilcox_default","wilcox_no_filter","hvg")
```
```{r}
max_logfc = read.table(file.path(folder_path, paste0(tissue_name,"_ground_truth_c1.txt"))) %>% rownames()
celldb_marker = read.table(file.path(folder_path, paste0(tissue_name,"_ground_truth_c2.txt")))[,1]
gt_list = c(lapply(seq(50,1000,50),function(x) max_logfc[1:x]),list(celldb_marker))
names(gt_list) = c(paste0("Top",seq(50,1000,50)),"CellMarkerDB")

@@

  df = roc_df %>% select(c("Method","AUROC")) %>% distinct()
  df$'gt_set' = names(gt_list)[i]
  df
}))
df$Method = as.factor(df$Method)
df$gt_set = factor(df$gt_set,levels = names(gt_list))
write.table(df,file = file.path(folder_path, paste0(tissue_name,"_auroc.csv")),row.names = FALSE)
```

@@

# Density Index
```{r}
df_benchmark = read.table(file.path(folder_path, paste0(tissue_name,"_benchmark_rank_table.csv")))

@@

write.table(score_measure,file = file.path(folder_path, paste0(tissue_name,"_density_index.csv")),row.names = FALSE)
```

# Sensitivity Test (Try different K)
```{r}
# RunLMD under differen K
source("./LMD_function.R", echo = F)
score_df = do.call(cbind,lapply(c(3:10,20,30,40,50),function(kNN){
  try({
    res = LMD(dat,feature_space,knn = kNN)
    return(res$cumulative_score)
  },silent = TRUE)
}))
colnames(score_df) = paste0("kNN",c(3:10,20,30,40,50))
write.table(score_df,file = file.path(folder_path, paste0(tissue_name,"_different_knn.csv")),row.names = TRUE)
```
```{r}
# Obtain AUROC
score_ranked <- data.frame(score_df) %>% mutate(across(everything(), rank))

max_logfc = read.table(file.path(folder_path, paste0(tissue_name,"_ground_truth_c1.txt"))) %>% rownames()
celldb_marker = read.table(file.path(folder_path, paste0(tissue_name,"_ground_truth_c2.txt")))[,1]

@@

auc_df = apply(score_ranked,2,function(x){
  unlist(lapply(gt_list,function(genes){
    roc = pROC::roc(as.integer(rownames(score_ranked) %in% genes), x, direction = ">")
    pROC::auc(roc)
  }))
})

auc_df = data.frame(auc_df,gt = names(gt_list))
write.table(auc_df,file = file.path(folder_path, paste0(tissue_name,"_sensitivity_auroc.csv")),row.names = FALSE)
```
